---
title: "Project_1_Analyst"
author: "go001"
date: "2/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
    library(dplyr)
    library(tidyr)
    library(tidyverse)
```

```{r Downloading Sample Data}
    matrix_file_path = "~/Desktop/batch_corrected_data.csv"
    matrix_data = read_delim(matrix_file_path, delim = ",")
    str(matrix_data)
```

#### Matrix Description
first column contains the probset ID and the other columns contain the sample information

```{r controlling the name}
    colnames(matrix_data) = sub("_.*", "", colnames(matrix_data))
    colnames(matrix_data)[1] = "Gene"
```
Since the probeset-id is currently labeled as pb-id, the column was renamed to be easily identifiable.


# Problem 4
Implement the following filters on the RMA normalized, ComBat adjusted expression matrix:

1. Expressed in at least 20% of samples (i.e. for each gene, at least 20% of the gene-expression values must be > 𝑙𝑜𝑔2(15)).
```{r filter for gene expression value}

# filters out the genes that do not have a high enough expression
min_level_expression = log(15,base = 2)
min_sample_expression = .2 
matrix_dataF1 =matrix_data %>% 
        filter( (rowSums(.[-1] > min_level_expression)/(ncol(.)-1)) >min_sample_expression )
```

```{r}
# Total amount of genes that had atleast 20% of the samples pass the gene expression values of atleast log(15,base = 2)
print(paste0("There was ", nrow(MaskF1)," genes that past the first filter."))
```


2. Have a variance significantly different from the median variance of all probe sets using a threshold of 𝑝<0.01 (hint: use a chi-squared test as shown in http://www.itl.nist.gov/div898/handbook/eda/section3/eda358.htm. 

You will need to manually compute the test statistic for each gene and compare it to the chi-squared distribution with the correct number of degrees of freedom using the qchisq() function).


```{r filter 2 significantly different variance}
#Function for obtaining the median variance
variance = function(x){
    a = sum((x - mean(x))^2)/(length(x) - 1)
    return(a)
}
# Variance for each row
# variance of every single gene
varList = apply(matrix_dataF1[,-1], MARGIN = 1, FUN = variance)
# Finding the median for the varList
sigma02 = median(varList)

# calculate the minimum test statistic 
testStat = qchisq(p = 0.01, df = ncol(matrix_data)-2,lower.tail = F)
df = ncol(matrix_data) - 2
testchi = df*varList/(sigma02)

# Create Mask for the second filter
MaskF2 = testchi>testStat

ptable = 1 - pchisq(testchi,ncol(matrix_data)-2)
matrix_dataF2 = matrix_dataF1[MaskF2,]
```

```{r}
# Total amount of genes that had atleast 20% of the samples pass the gene expression values of atleast log(15,base = 2)
print(paste0("There was ", nrow(matrix_dataF2)," genes that past the first filter."))
```

Have a coefficient of variation > 0.186.

```{r filter 3 coefficient of variation}
    varListF3 = varList[MaskF2]
    mu = apply(matrix_dataF2[,-1], MARGIN = 1, FUN = mean)
    MaskF3 = (sqrt(varListF3)/mu) > 0.186
    matrix_dataF3 = matrix_dataF2[MaskF3,]
```


3. Write out a different file containing the gene expression matrix for genes passing all three of the filters from 4.1, 4.2, and 4.3.

```{r}
write.csv(matrix_dataF3,file = "./matrix_data_3_filter.csv",row.names = F)
```


4. For groups with Biologist role only: Write out the expression matrix for probesets that pass the expression threshold from 4.2 to a file with write.csv.

```{r}
write.csv(write_Bio, file = "./write.csv", row.names = F)
```

Part 5.

```{r Setting up data for clustering }
# clustering will be done on the matrix that passed all 3 filters
clustering_data = matrix_dataF3

# Currently the rows are the probesets so the clustering will be done on the probesets instead of samples
# the matrix is Transposed
TcolName = matrix_dataF3$Gene
TrowName = names(matrix_dataF3)[-1]
df.clustering_data = as.data.frame(t(clustering_data[,-1]))

# Its important to keep track of the probeset ids and sample ids so it is placed back in 
names(df.clustering_data) = TcolName
row.names(df.clustering_data) = TrowName
#df.clustering_data
```

```{r Clustering}
# this is the clustering using euclidean distance
dist_mat <- dist(df.clustering_data, method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg,cex=0.8)
```

```{r Split into two clusters}

cut_avg <- cutree(hclust_avg, k = 2)

plot(hclust_avg, cex=0.8)

suppressPackageStartupMessages(library(dendextend))
avg_dend_obj <- as.dendrogram(hclust_avg)
avg_col_dend <- color_branches(avg_dend_obj, h = 90)
plot(avg_col_dend)
rect.hclust(hclust_avg , k = 2, border = 2:6, cex=0.7)
```

```{r Counting the amount of samples per cluster}
    k = 2
    for(i in 1:k){
        print(paste0("There is ", sum(cut_avg==i), " samples in cluster ",i))
    }
```

```{r Creating Heat Map}
#str(df.clustering_data)
#col contains samples
#row contains genes
#df.clustering_data

# Reads the metadata and record which Geo accesion id(Sample ID) is associated with which colon subtype
meta_data = read_delim("~/Desktop/proj_metadata.csv", delim = ",")
geo_accession = rownames(df.clustering_data)
rowName = as.data.frame(geo_accession)
classification = dplyr::left_join(x = rowName, y = meta_data[c("cit-coloncancermolecularsubtype","geo_accession")], by= "geo_accession")

# Creates a mask to determine which colon subtype is associated with which geo_accesion
C3_subtype = classification$`cit-coloncancermolecularsubtype`=="C3"
C3_subtype= C3_subtype[1:nrow(df.clustering_data)]
data_heatmap <- as.matrix(t(df.clustering_data))

# Create with Heatmap
png(file="~/Desktop/heatmap.png",width = 1000,height = 800)
heatmap(x = data_heatmap,ColSideColors = if_else(C3_subtype,"Red","blue"),cexCol = 0.8, scale = "row")
legend(x="right", legend=c("1:Low", "2","3","4", "5:High"),fill=heat.colors(5),cex = 0.8,title = "Heat map")
legend(x="topright", legend=c("C3", "Other"),fill=c("Red","Blue"),cex = 0.8,title = "Colon Cancer Subtype")

```

```{r}
#attempting Welch test
write_Bio = data_frame(Gene = c(),t_statistic = c(),p_value = c(), p_adj = c())
for( i in 1:ncol(df.clustering_data)){
    t_test = t.test(df.clustering_data[C3_subtype,i], df.clustering_data[-C3_subtype,i], alternative = "two.sided")
    p_adj = p.adjust(t_test$p.value, method = "bonferroni", n = ncol(df.clustering_data))
    temp = data_frame(Gene = names(df.clustering_data)[i], t_statistic = t_test$statistic,p_value = t_test$p.value, p_adj = p_adj)
    write_Bio = rbind(write_Bio,temp)
}

End_of_Analyst = data_frame(write_Bio$Gene,write_Bio$t_statistic,write_Bio$p_value,write_Bio$p_adj)
names(End_of_Analyst) = c("Gene", "t_score", "p_value", "p_adj")
write.csv(End_of_Analyst, file = "./write.csv", row.names = F)
```
```{r}
# displaying the result of the expression types
Highly_expressed = End_of_Analyst %>%
    filter(p_adj<0.05) %>%
    arrange(p_adj)

print(paste0("There was ", sum(Highly_expressed$t_score>0)," genes up-regulated."))
print(paste0("There was ", sum(Highly_expressed$t_score<0)," genes down-regulated."))

Highly_expressed[Highly_expressed$t_score>0,]
```


```{r for Biologist}
clustering_data = matrix_dataF2
TcolName = matrix_dataF2$Gene
TrowName = names(matrix_dataF2)[-1]

df.clustering_data = as.data.frame(t(clustering_data[,-1]))
names(df.clustering_data) = TcolName
row.names(df.clustering_data) = TrowName

write_Bio = data_frame(Gene = c(),t_statistic = c(),p_value = c(), p_adj = c())
for( i in 1:ncol(df.clustering_data)){
    t_test = t.test(df.clustering_data[C3_subtype,i], df.clustering_data[-C3_subtype,i], alternative = "two.sided")
    p_adj = p.adjust(t_test$p.value, method = "bonferroni", n = ncol(df.clustering_data))
    temp = data_frame(Gene = names(df.clustering_data)[i], t_statistic = t_test$statistic,p_value = t_test$p.value, p_adj = p_adj)
    write_Bio = rbind(write_Bio,temp)
}

For_Bio = data_frame(write_Bio$Gene,write_Bio$t_score,write_Bio$p_value,write_Bio$p_adj)
write.csv(write_Bio, file = "./write.csv", row.names = F)
```
